---
title: "Reproducible Research: personal activity monitoring device"
author: "S. Pribis"
date: "Tuesday, October 07, 2014"
output: html_document
---
## Loading and preprocessing the data
### Loading the data
```{r load_data}
fileURL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileURL, "repdata-data-activity.zip", mode = "wb")
data <- read.csv(unz("repdata-data-activity.zip", "activity.csv"))
```

### Processing/transforming the data
```{r process_data}
library(reshape2)
data_melt <- melt(data, id=c("date"), measure.vars="steps")
data_sum <- dcast(data_melt, date ~ variable, sum)
data_melt2 <- melt(data, id=c("interval"), measure.vars="steps")
data_mean2 <- dcast(data_melt2, interval ~ variable, mean, na.rm=T)
data_median2 <- dcast(data_melt2, interval ~ variable, median, na.rm=T)
```

## What is mean total number of steps taken per day?
```{r total_steps_day}
with(data_sum, hist(steps, main = "histogram of total steps per day"))
options(scipen=999)
day_mean <- with(data_sum, round(mean(steps, na.rm=T)))
day_median <- with(data_sum, round(median(steps, na.rm=T)))
```
The mean and median steps per day are `r day_mean` and `r day_median`, respectively.

## What is the average daily activity pattern?
```{r daily_activity_pattern}
with(data_mean2, plot(steps ~ interval, type="l", main = "average steps per interval by interval"))
max_steps_interval <- data_mean2$interval[order(data_mean2$steps, decreasing = T)[1]]
```
The interval with the largest average steps per day across all days is `r max_steps_interval`.

## Imputing missing values
```{r NAs}
missing_values <- sum(complete.cases(data)==F)
```

There are `r missing_values` rows with missing values in the dataset.  
The `r NA` values will replaced by the median value for that interval across all days.

```{r new_data}
new_data <- data
for(i in as.integer(row.names(data[(complete.cases(data)==F),]))) {
        new_data[i, "steps"] <- merge(x=data[i,], y=data_median2, 
                                      by="interval")[1,"steps.y"]
}
```

```{r new_hist}
new_data_melt <- melt(new_data, id=c("date"), measure.vars="steps")
new_data_sum <- dcast(new_data_melt, date ~ variable, sum)
with(new_data_sum, hist(steps, main = "histogram of total steps per day (imputed data)"))

new_day_mean <- with(new_data_sum, round(mean(steps, na.rm=T)))
new_day_median <- with(new_data_sum, round(median(steps, na.rm=T)))

mean_diff <- day_mean - new_day_mean
median_diff <- day_median - new_day_median

perc_change_mean <- round(mean_diff*100/day_mean,1)
perc_change_median <- round(median_diff*100/day_median,1)
```

After replacing the `r NA` values, the mean and median steps per day are `r new_day_mean` and `r new_day_median`, respectively.  
  
The histogram displays the impact of imputing missing data on the distribution of daily steps. All 8 days previously missing data now fall into the bin with the least number of steps. This greatly affects the mean number of steps, which decreased `r perc_change_mean`% or `r mean_diff` steps. The effect on the median was not as significant; it decreased `r perc_change_median`% or `r median_diff` steps.

## Are there differences in activity patterns between weekdays and weekends?
```{r weekends}
data$Date <- as.Date(data$date,"%Y-%m-%d")

for (i in 1:length(data$Date)) {
        data$day[i] <- if(weekdays(data$Date[i])=="Sunday" | 
                                  weekdays(data$Date[i])=="Saturday") {
        "weekend"
        } else {"weekday"}
}

w_data_melt2 <- melt(data, id=c("interval", "day"), measure.vars="steps")
w_data_mean2 <- dcast(w_data_melt2, interval + day ~ variable, mean, na.rm=T)

library(ggplot2)
g <- ggplot(w_data_mean2, aes(interval, steps))
g + geom_line() + facet_grid(day ~ .) + labs(title = "differences in activity patterns between weekdays and weekends")
```